---
title: "rasterPrep preparing raster files for modeling or viewing"
author: "Ethan Plunkett"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rasterPrep}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# rasterPrep 
The rasterPrep package provides a workflow to prepare data for modeling by creating a standardized set of rasters from disparate raster and vector GIS data; and a second workflow for  preparing raster files for viewing in GIS software.  These workflows are each documented in their own section below.

### Installation

This package requires that GDAL Utilities be installed on your computer and the directory that houses them be added to your PATH environmental variable.  On Linux and Mac systems I recommend you use your package manager to install GDAL.  On Windows I recommend you download the latest "Stable release" from http://download.gisinternals.com/; you want the "Generic installer for the GDAL core components".   With considerably more effort you can install GDAL Utilities on Windows with python bindings but that is not required for this package;  If you want to anyway read [this](https://sandbox.idre.ucla.edu/sandbox/tutorials/installing-gdal-for-windows).


## I. Preparing rasters for modeling

This section walks through the process of preparing raster data for modeling.

Often the first step in any raster based GIS modeling exercise is creating a series of raster files that coincide perfectly. That is they agree in cell size, cell alignment, projection, and extent. Typically, the source data is a mix of raster and vector files that do not coincide.  This workflow demonstrates how to fix that.

Overview:

1. Load packages and setup paths
2. Create a reference raster file
3. Use `warpToReference()` to make existing rasters coincide with the reference
4. Use `rasterizeToReference()` to convert vector data to rasters that coincide with the reference.

### 1. Load packages and setup paths
```{r, echo = TRUE, message = FALSE, warning = FALSE}
# Load packages
library(rasterPrep)
library(gdalUtilities)  # for ogr2ogr() used to reproject vector Shapefiles
library(terra)  # for crs() and rast() 

# Note because I'm pulling data from one read-only directory and writing to a
# a second I'm defining complete paths here.  Often it's easier to set a working
# directory then use file names when calling the functions.

# Set paths for retrieving input data included with this package
inPaths <- list(
  bound = system.file("extdata","Amherst.shp" , package = "rasterPrep"),
  slope = system.file("extdata","slope.tif" , package = "rasterPrep"), 
  roads =  system.file("extdata","roads.shp" , package = "rasterPrep"),
  key = system.file("extdata","roadClassKey.csv" , package = "rasterPrep")
)

# Output directory: you might want to replace this with a path of your choosing:
outDir <- tempdir() 

# Set output paths 
outPaths <- list(
  ref = file.path(outDir, "reference.tif"),        # Our reference grid
  slopeTemp = file.path(outDir, "slopeTemp.tif"),  # Intermediate raster file
  slope = file.path(outDir, "slope.tif"),          # Slope raster
  roadLines = file.path(outDir, "roadLines.shp"), # Reprojected vector roads
  roadsTemp = file.path(outDir, "roads1.tif"),     # Intermediate raster file
  roads  = file.path(outDir, "roads.tif"),         # Binary Roads raster
  roadClass  = file.path(outDir, "roadClass.tif") )# Road class raster

# Delete output so rerunning doesn't cause problems
if(file.exists(outPaths$roadLines))
  file.remove(list.files(outDir, "roadLines", full.names = TRUE))

for(a in setdiff(names(outPaths), "roadLines")){
  if(file.exists(outPaths[[a]]))
    deleteTif(outPaths[[a]])
}


```

The four  data sets included with this package are:

1.  "Amherst.shp":  The polygon boundary for the Town of Amherst, MA extracted from here:
https://docs.digital.mass.gov/dataset/massgis-data-community-boundaries-towns
It is in Mass Mainland mass state plain projection  ([EPSG:26986](https://epsg.io/26986)).
This defines our study area and output projection.

2. "slope.tif": A raster clipped from https://scholarworks.umass.edu/data/17/ with 
the original projection and cell alignment preserved.  It is in the USA Contiguous Albers Equal Area Conic Projection([ESRI:202003](https://epsg.io/102003)).

3. "roads.shp": A Shapefile containing vector roads, originally from Open Street map but with small roads removed and clipped to Amherst.  It's in the same projection as 2. 

4. "roadClassKey.csv": A CSV file with columns "value" , "category", and "color", representing integer values, associated road class names, and a hexadecimal color that should be used for rendering (#rrggbb).

### 2. Create a reference raster

You may already have a raster that you would like all your other data to coincide with.  If so use that as your reference.  If you do not already have a reference raster then we can create one from a polygon that delineates a study area boundary and has the projection we would like to work in.
```{r, echo = TRUE, message = FALSE, warning = FALSE, results = FALSE}

makeReference(polyFile = inPaths$bound, destination = outPaths$ref, cellsize = 30)

```

Note: The optional 'alignTo' argument determines the cell alignment (AKA "snapping"); the default (`"origin"`) means that the edges of the cells will all fall on integer multiples of the cell size. If you already have some raster data in your desired projection you can also set alignTo="reference" and then set the reference  argument to the path of a raster that will set the cell alignment.  In this case the reference would only be used for cell alignment and not extent; allowing you to use that data without resampling or shifting the pixels. With either pixel alignment the resulting raster is going to have an extent just big enough to include all the pixels that overlap the bounds of the study area as defined in polyFile. 

### 3. Warp raster data to match the reference

Here we will align the sample slope data with our reference raster.  We will reproject 
and crop in one step and then clip (assign NA outside of the boundary polygon) 
in a second step. I've found that although we can combine these all in one step
the masking of pixels outside of the boundary isn't always consistent on the edges when reprojection and clipping are done together.

If you don't want to clip you could just use the first step and you would retain data
for the entire rectangular extent.

Because slope is continuous we'll use the `"bilinear"` resampling method 
in the first step.  If we were working with categorical data (E.g. land cover
classes) we would leave the default nearest neighbor (`"near"`) method.


```{r, echo = TRUE, message = FALSE, warning = FALSE, results = FALSE}

# Reproject to final extent, projection and alignment
warpToReference(inPaths$slope, outPaths$slopeTemp, reference = outPaths$ref, method = "bilinear")

# Assign NA outside of boundary
warpToReference(outPaths$slopeTemp, outPaths$slope, reference =  outPaths$ref, clip = inPaths$bound)

# Delete intermediate file
deleteTif(outPaths$slopeTemp)

```

### 4. Rasterize vector to match the reference

Here we will first reproject the vector roads to our output projection and then 
use the result to create a roads raster that matches the reference raster.

To reproject the vector roads we are going to use the 'ogr2ogr()' function within the
gdalUtilities R package.  You could also use sf package to read in the vector data, 
reproject, and write out to a new Shapefile, but it woud be less efficient.

```{r, echo = TRUE, message = FALSE, warning = FALSE, results = FALSE}

# Reproject vector to reference projection
refProj <- terra::crs(terra::rast(outPaths$ref))  #  well known text of projection
projFile <- file.path(outDir, "refproj.txt")
cat(refProj, file = projFile)   #well known text in file
gdalUtilities::ogr2ogr(inPaths$roads, outPaths$roadLines, t_srs = projFile)

```

Now we are going to rasterize the vector data using `rasterizeToReference()` from this package.
We could do this in several ways depending on what data we want in the final
raster. Importantly this function will write into any existing destination raster files
so multiple calls can be used to overlay data from different sources together
in a file.

In this first example we'll create a raster that contains the values from the 
"ROADCLASS"" attribute from the roads shapefile in cells underlying the roads 
and and NA everywhere else.

```{r, echo = TRUE, message = FALSE, warning = FALSE, results = FALSE}
rasterizeToReference(outPaths$roadLines, destination = outPaths$roadClass, reference = outPaths$ref, attribute = "ROADCLASS", type = "byte")

```

Next an example where we'll create a raster with NA outside the study area 
boundary,  and 0 or 1 inside the boundary to indicate if there is a road. In this
example we'll also set `allTouched=TRUE` which will result in roads that have
orthogonal connections.

```{r, echo = TRUE, message = FALSE, warning = FALSE, results = FALSE}

# Burn in zero everywhere within the study area boundary creating a new raster file in the process
rasterizeToReference(inPaths$bound, outPaths$roadsTemp, 
                     reference = outPaths$ref, burn = 0, allTouched = TRUE)

# Burn in 1 where there are roads (on top of values from prior step)
rasterizeToReference(outPaths$roadLines, destination = outPaths$roadsTemp, 
                     burn = 1)

# Write to new file while clipping.  Without this step road cells outside of the 
#  study boundary will have a value of 1. This re-sets them to NA.
warpToReference(outPaths$roadsTemp, outPaths$roads, reference = outPaths$ref,
                clip = inPaths$bound)

# Delete intermediate file
deleteTif(outPaths$roadsTemp)
```

Print out the paths to what we created:
```{r, results= FALSE}
cat("Files created here:\n\t",
    normalizePath(outPaths$ref), " our reference raster\n\t",
    normalizePath(outPaths$roadClass), " road class\n\t", 
    normalizePath(outPaths$roads), " roads 1, other study area cells 0\n\t", 
    normalizePath(outPaths$slope), " slope."
    )

# Confirm that projections information is there for all files
# Not required but changes in R's handling of spatial data have broken this in the past
refcrs <- terra::crs(terra::rast(outPaths$ref))  # wkt of refrence grid we created
if(refcrs == "") stop("Reference grid doesn't have projection information")
for(file in c("slope", "roads", "roadClass")){
  filecrs <- terra::crs(terra::rast(outPaths[[file]]))
  stopifnot(filecrs == refcrs)
}

```


## II. Preparing raster files for viewing with GIS software

The workflow in this section is for creating .tif files that are ideally suited for viewing with GIS software. These are compressed, tiled, and have overviews (AKA pyramids).  Depending on the data contained in the file we may also add statistics, a color table, and/or a Value Attribute Table (VAT).  Tiling and overviews allow GIS software to efficiently display the data across a wide range of scales and compression reduces file size. 

Everything we do here either hinders reading and processing with the raster package or adds extra files that don't help, so I do not recommend running these functions on the files you plan
to process with R. I use these to prep output rasters for viewing and distribution.

Tiling and overviews aren't necessary for the small demonstration files we are using but make a huge difference with large datasets.  

For many datasets a single function call will create a copy of the input raster that is optimized for viewing. Here's an example with the slope grid produced above. It is a continuous grid so I'm setting `overviewResample="average"` so that each overview cell will be assigned the average value of all the original cells it covers.

```{r, results= FALSE}
  final.dir <- file.path(outDir, "final")
 dir.create(final.dir, showWarnings = FALSE)
 makeNiceTif(outPaths$slope, file.path(final.dir,  "slope.tif") , overviewResample = "average")

```

For categorical data we may want to add a VAT and color table. I've found its very difficult to embed color tables in a tif file using R. The method used by `addColorTable()` works only for byte encoded grids; it also requires every integer between 0 and the maximum value in the grid to appear in the color table. Finally, it is displayed in a somewhat clunky way in ArcGIS.  However, it is best way I've found to programmatically set the color and labels for categories in a tif file from within R; If you have a better way please let me know.

We'll demonstrate with the road class grid we created above.  First, we'll use `addColorTable()` to create a small .vrt (gdal virtual format) file that references the tif and adds color table information. 
The .vrt file can be opened directly with ArcGIS or QGIS but we'll use it to create a .tif with the color table embedded.


```{r results= FALSE}
key <- read.csv(inPaths$key, stringsAsFactors = FALSE)  

# Note the table must have "value", and "color" columns and can optionally 
# have a "category" column will class labels
vrt.file <- addColorTable(outPaths$roadClass, table = key)

```

Next well encode the data and the color table into a new .tif. along with compression,
tiling, and overviews.  The default `overviewResample` of `"near"` is appropriate for
categorical data.
```{r}
finalRoadClass <- file.path(final.dir,  "roadClass.tif")
makeNiceTif(source = vrt.file, destination = finalRoadClass, overwrite = TRUE, stats = FALSE )
```

Finally, we can optionally add an ESRI style value attribute table (VAT) sidecar file to the dataset, that includes the attribute information from the key. 

```{r}

addVat(x = finalRoadClass, attributes = key)

```


If you use ArcGIS you may want to add a VAT to a categorical raster even if you don't want to add attribute information or a color table.  The presence of a VAT causes ArcGIS to default to a unique value display of the data and saves the user from having to wait in an interactive session for ArcGIS to build the VAT when you switch symbology to categories. In this case we can do everything in a single call to the `makeNiceTif()` function:
```{r}
roadWithVAT <- file.path(final.dir,  "roadsWithVAT.tif")
makeNiceTif(source = outPaths$roadClass, destination = roadWithVAT, overwrite = TRUE, overviewResample = "near", stats = FALSE, vat = TRUE )
```



